---
title: "Analysis and figures"
author: "Saoirse Kelleher"
date: today
abstract: "Calculates key statistics and renders figures for \"*Twenty years of dynamic occupancy models: a review of applications and look to the future*\""
abstract-title: "Summary"
format:
  html:
    theme: minty
---

## Load packages

First, we load in a handful of packages required for analysis and visualisation.

```{r Load packages}
#| message: false

library(tidyverse)
library(patchwork)
library(sf)
library(readxl)
library(ggtext)
```

## Read in data

Two data sources are required to render the analyses and visualisation for this article:

i.  Randomised list of all queried articles *(Randomisation.xlsx)*

ii. Spreadsheet of articles included in the review *(Review_Spreadsheet.xlsx)*

#### Randomised article list

All queried articles were collated and assigned random values in *'Randomisation.qmd'* to determine the order in which they would be considered for inclusion. '*Randomisation.xlsx'* contains all of these articles, and also documents which articles were included in the review, were excluded based on our review criteria, or were not processed at all. This spreadsheet is split across sheets by year strata. All sheets are read in using `readxl` and combined.

```{r Load query sheets}
sheet_AllArticles <- list(read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2004-2007"),
                          read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2008-2011"),
                          read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2012-2015"),
                          read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2016-2019"),
                          read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2020-2023")) |> 
  list_rbind()
```

#### Reviewed data

*'Review_Spreadsheet.xlsx'* contains all recorded details from the 92 reviewed articles. This spreadsheet contains several sheets, each including one category of variable (e.g., details on 'Focal Taxa' or 'Covariates'). Each of these is loaded individually with `readxl`.

```{r Load review sheets}
sheet_ArticleData <- read_xlsx("Review_Spreadsheet.xlsx",
                               sheet = "Article Data")
sheet_Objectives <- read_xlsx("Review_Spreadsheet.xlsx",
                              sheet = "Objectives") 
sheet_FocalTaxa <- read_xlsx("Review_Spreadsheet.xlsx",
                             sheet = "Focal Taxa") 
sheet_StudyArea <- read_xlsx("Review_Spreadsheet.xlsx",
                             sheet = "Study Area")
sheet_DataCollection <- read_xlsx("Review_Spreadsheet.xlsx",
                                  sheet = "Data Collection")
sheet_Covariates <- read_xlsx("Review_Spreadsheet.xlsx",
                              sheet = "Covariates")
sheet_Modelling <- read_xlsx("Review_Spreadsheet.xlsx",
                             sheet = "Modelling")
```

## Article inclusion

The first set of numbers we calculate describe the articles included in the review, and the inclusion rates within each of our year strata. These values are used to render a figure showing the coverage of reviewed articles during our study period.

#### Articles by query

Recall that our review sample was drawn from two Web of Science queries: the first being all articles which have cited @mackenzie2003, and the second using a selection of keywords presumed to be used in articles fitting DOMs. Here, we calculate the number of articles which appeared in one or both of these queries.

```{r Calculate query statistics}
stat_queryCounts <- sheet_AllArticles |>
  mutate(Query = case_when(KeyTerms == TRUE & CitesMackenzie == TRUE ~ "Both",
                           KeyTerms == FALSE & CitesMackenzie == TRUE ~ "Mackenzie only",
                           KeyTerms == TRUE & CitesMackenzie == FALSE ~ "Key terms only")) |>
  summarise(Articles = n(), .by = Query) |>
  mutate(Proportion = round(Articles/sum(Articles), 2))
stat_queryCounts
```

#### Total articles

Next, we get the number of articles included in the review.

```{r Total article count}
stat_articleTotal <- sheet_AllArticles |>
  filter(Included == "YES") |>
  nrow()
stat_articleTotal
```

#### Article hit rates

We then calculate the hit rate for articles in each stratum, by dividing the number of articles included by the number of articles processed. These hit rates are then used to estimate how many articles from the remaining pool would have met inclusion criteria — first summed from each year, then summed across years.

```{r Calculate article hit rates}
# Hit rates within each stratum
stat_hitRates <- sheet_AllArticles |>
  summarise(Articles = n(), 
            .by = c(Year, Strata, Included)) |>
  pivot_wider(names_from = Included, 
              values_from = Articles, 
              values_fill = 0) |>
  summarise(HitRate = round(sum(YES)/(sum(NO)+sum(YES)), 2),
            .by = Strata)
stat_hitRates

# Number of remaining articles per year
stat_remainingArticles <- sheet_AllArticles |>
  summarise(Articles = n(), .by = c(Year, Strata, Included)) |>
  pivot_wider(names_from = Included, 
              values_from = Articles, values_fill = 0) |>
  mutate(HitRate = sum(YES)/(sum(NO)+sum(YES)),
         .by = Strata) |>
  mutate(EstRemaining = round(`NA`*HitRate, 2)) |>
  select(Year, EstRemaining)
stat_remainingArticles

# Estimating remaining qualified articles and total remaining articles
stat_remainingProportion <- data.frame(Group = c("EstRemaining", "TotalUnreviewed"),
                                  Articles = c(sum(stat_remainingArticles$EstRemaining),
                                               sheet_AllArticles |>
                                                 summarise(Articles = n(), 
                                                           .by = c(Year, 
                                                                   Strata, 
                                                                   Included)) |>
                                                 pivot_wider(names_from = Included, 
                                                             values_from = Articles,
                                                             values_fill = 0) |>
                                                 pull(`NA`) |> 
                                                 sum()))
stat_remainingProportion
```

#### Coverage figure

Using these values we render a figure presenting the estimated coverage of our review over time (Figure 3 in the article).

```{r Coverage figure}
#| fig-width: 8
#| fig-height: 4

plot_reviewCoverage <- 
  sheet_AllArticles |>
  summarise(N = n(), .by = c(Year, Strata, Included)) |>
  pivot_wider(names_from = Included, values_from = N, values_fill = 0) |>
  mutate(HitRate = sum(YES)/(sum(NO)+sum(YES)),
         .by = Strata) |>
  mutate(EstRemaining = `NA`*HitRate) |>
  pivot_longer(cols = c(YES, EstRemaining), 
               names_to = 'Group', values_to = 'Articles') |>
  
  ggplot() +
  geom_col(aes(x = Year, y = Articles, fill = Group), colour = "white") +
  scale_fill_manual("",
                    limits = c("YES", "EstRemaining"),
                    labels = c("Included articles", 
                               "Estimated qualifying articles"),
                    values = c("aquamarine4", "gray65")) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 65)) +
  labs(x = "", y = "Number of articles") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "bottom", 
        axis.text = element_text(size = 12),
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title = element_text(size = 12), 
        legend.text = element_text(size = 12),
        strip.text.x = element_text(size = 12)) +
  facet_grid(cols = vars(Strata), scales = "free_x", switch = "x")
plot_reviewCoverage
ggsave(plot = plot_reviewCoverage, "Figures/CoveragePlot.jpeg",
       width = 8, height = 4)
```

## Study areas

The next set of statistics and figures focus on the study areas included in reviewed articles. For each article the countries where data was collected, a rough centroid of the study area, and an estimate of the spatial extent of the study area was recorded.

#### Study countries table

First, a table is generated presenting the number of articles using data from each country with at least one article.

```{r Study location table}
stat_studyCountries <- sheet_StudyArea |>
  select(`Review ID`, Country) |>
  separate_longer_delim(Country, ", ") |>
  distinct() |>
  mutate(TotalArticles = length(unique(`Review ID`))) |>
  summarise(Articles = n(), .by = c(Country, TotalArticles)) |>
  mutate(Proportion = round(Articles/TotalArticles,2)) |>
  select(-TotalArticles) |>
  arrange(-Articles)
stat_studyCountries
```

#### Unique countries

The total number of unique countries is extracted.

```{r Unique countries}
stat_uniqueCountries <- stat_studyCountries$Country |>
  unique() |>
  length()
stat_uniqueCountries
```

#### Map of study locations

A figure is rendered plotting the rough centroids for each study area included in the review.

::: callout-important
## Some articles include multiple study areas which were treated as distinct when they were modelled independently.
:::

```{r Study location map}
# Load country shapefile
countries_shp <- st_read("Analysis/Countries_Spatial/World_Countries__Generalized_.shp", quiet = TRUE) |>
  st_transform(crs = "ESRI:54030") |>
  filter(COUNTRY != "Antarctica")

studyArea_Map <- sheet_StudyArea |>
  st_as_sf(coords = c("Centre Point (lon)", "Centre Point (lat)"),
           crs = "EPSG:4326") |>
  st_transform(crs = "ESRI:54030") |>
  
  ggplot() +
  geom_sf(data = countries_shp, colour = "gray90", fill = "gray80") +
  geom_sf(shape = 21, fill = "aquamarine4", colour = "white", alpha = 0.5, size = 3.5) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
studyArea_Map
ggsave(plot = studyArea_Map, "Figures/StudiesMap.png")
```

#### Study area size

```{r Study area size}
#| fig-width: 7
#| fig-height: 4

studyAreaPlot <- sheet_StudyArea |>
  summarise(Articles = n(), .by = Size) |>
  filter(Size != "Unspecified") |>
  
  ggplot() +
  geom_col(aes(x = Size, y = Articles),
           fill = "aquamarine4", colour = "white",
           width = 1, linewidth = 2) +
  labs(x = "", y = "") +
  scale_x_discrete(labels = c("< 1", "1 - 10", "10 - 10<sup>2</sup>",
                              "10<sup>2</sup> - 10<sup>3</sup>",
                              "10<sup>3</sup> - 10<sup>4</sup>",
                              "10<sup>4</sup> - 10<sup>5</sup>",
                              "10<sup>5</sup> - 10<sup>6</sup>",
                              "\\> 10<sup>6</sup>")) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_markdown(size = 20, angle = 90,
                                       vjust = 0.5, hjust = 1),
        axis.text.y = element_blank(), axis.ticks = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA))
studyAreaPlot

ggsave("Figures/StudyAreaPlot.png",
       studyAreaPlot, width = 7, height = 4, bg='transparent')
```

## Study taxa

#### Multi-species statistics

```{r Multi-species stats}
sheet_FocalTaxa |>
  separate_longer_delim(`Multispecies method`, ", ") |>
  select(`Review ID`, `Multispecies method`) |>
  distinct() |>
  mutate(TotalArticles = length(unique(`Review ID`))) |>
  summarise(Articles = n(), 
            .by = c(`Multispecies method`, TotalArticles)) |>
  mutate(Proportion = round(Articles/TotalArticles, 2))
```

#### Study species

```{r Focal species figure}
#| fig-width: 10
#| fig-height: 4

focalTaxaPlot <- 
  sheet_FocalTaxa |>
  separate_longer_delim(`Taxa keywords`, delim = ", ") |>
  separate_longer_delim(Status, delim = ", ") |>
  mutate(`Taxa keywords` = case_when(`Taxa keywords` == "Fish" ~ "Other", 
                                     .default = `Taxa keywords`)) |>
  mutate(Status = case_when(Status %in% c("CR", "EN", "NT", "VU", "Threatened_Other") ~ "Threatened",
                            Status %in% c("NA", "Unclear", "DD") ~ "NA",
                            Status == "Invasive" ~ "Invasive",
                            Status == "LC" ~ "Stable")) |>
  mutate(Threatened = case_when("Threatened" %in% Status ~ TRUE,
                                .default = FALSE),
         Invasive = case_when("Invasive" %in% Status ~ TRUE,
                              .default = FALSE),
         Overall = TRUE,
         .by = c(`Review ID`, `Taxa keywords`)) |>
  select(`Review ID`, `Taxa keywords`, Threatened, Invasive, Overall) |>
  distinct() |>
  pivot_longer(cols = c(Threatened, Invasive, Overall), 
               names_to = "Group", values_to = "Value") |>
  summarise(Articles = sum(Value),
            .by = c(`Taxa keywords`, Group)) |>
  mutate(Group = fct(Group, levels = c("Overall", "Threatened",
                                       "Invasive"))) |>
  mutate(`Taxa keywords` = fct(`Taxa keywords`, 
                               levels = rev(c("Bird", "Mammal", "Herptile",
                                              "Invertebrate", "Other")))) |>
  mutate(Articles = case_when(Articles == 0 ~ 0.1, .default = Articles)) |>
  
  ggplot() +
  geom_col(aes(x = Articles, y = `Taxa keywords`, fill = Group),
           position = position_dodge2(reverse = TRUE), width = 0.8) +
  scale_fill_manual("", values = c("aquamarine4", "goldenrod2", "gray70")) +
  scale_x_continuous(expand = c(0,0)) +
  labs(x = "Articles", y = "") +
  theme(panel.grid = element_blank(),
        legend.text = element_text(size = 20), axis.text = element_text(size = 21),
        axis.title = element_text(size = 19), axis.ticks.y = element_blank(),
        panel.background = element_rect(fill='transparent', color = NA),
        plot.background = element_rect(fill='transparent', color = NA),
        legend.background = element_rect(fill = 'transparent', color = NA))
focalTaxaPlot

ggsave("Figures/TaxaPlot.png",
       focalTaxaPlot, width = 10, height = 4, bg='transparent')
```

## Data collection

#### Survey method statistics

```{r survey methods}
sheet_DataCollection |>
  separate_longer_delim(`Capture method`, ", ") |>
  select(`Review ID`, `Capture method`) |>
  distinct() |>
  mutate(TotalArticles = length(unique(`Review ID`))) |>
  summarise(Articles = n(), .by = c(`Capture method`, TotalArticles)) |>
  mutate(Proportion = round(Articles/TotalArticles, 2))

sheet_DataCollection |>
  select(`Review ID`, Citizen) |>
  distinct() |>
  mutate(Citizen = case_when(Citizen == "YES" ~ TRUE,
                             Citizen == "NO" ~ FALSE)) |>
  pull(Citizen) |>
  sum()
```

#### Site quantity

```{r site quantity plot}
#| fig-width: 12
#| fig-height: 3.5

siteQuantityPlot <- sheet_DataCollection |>
  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
  ggplot() +
  geom_histogram(aes(x = `Site quantity`),
                 binwidth = 0.25, fill = "aquamarine4", colour = "white",
                 linewidth = 2) +
  geom_vline(aes(xintercept = median(`Site quantity`, na.rm = TRUE)), 
             colour = "goldenrod2", linetype = 1, linewidth = 2) +
  scale_x_continuous(transform = "log10",
                     limits = c(5, 10000),
                     breaks = c(10, 25, 50, 100, 250, 500,
                                1000, 2500, 5000),
                     expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), position = "right") +
  labs(y = "Site\nquantity", x = "") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA),
        axis.text.x = element_text(size = 23, angle = 90,
                                   vjust = 0.5, hjust = 1),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_line(), 
        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
siteQuantityPlot
ggsave(plot = siteQuantityPlot, filename = "Figures/Sites.png",
       height = 3, width = 12)
```

```{r site quantity statistics}
sheet_DataCollection |>
  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
  summarise(Sites = mean(`Site quantity`, na.rm = TRUE), 
            .by = `Review ID`) |>
  pull(Sites) |>
  median(na.rm = TRUE)
```

#### Study duration

```{r site duration figure}
#| fig-width: 12
#| fig-height: 3.5

studyDurationPlot <-
  sheet_DataCollection |>
    mutate(`Start month` = case_when(str_detect(`Start month`, "[:digit:]{5}") ~ `Start month`,
                                     .default = NA),
           `End month` = case_when(str_detect(`End month`, "[:digit:]{5}") ~ `End month`,
                                     .default = NA)) |>
  mutate(StartDay = as.numeric(as.Date(as.numeric(`Start month`))),
         EndDay = as.numeric(as.Date(as.numeric(`End month`)))) |>
  mutate(StudyDuration = EndDay - StartDay) |>
  mutate(StudyMonths = (StudyDuration/30)) |>
  mutate(StudyMonths = case_when(StudyMonths == 0 ~ 1,
                                 .default = StudyMonths)) |>
  
  ggplot() +
  geom_histogram(aes(x = StudyMonths), 
                 binwidth = 0.2, fill = "aquamarine4", colour = "white",
                 linewidth = 2) +
  geom_vline(aes(xintercept = median(StudyMonths, na.rm = TRUE)), 
             colour = "goldenrod2", linetype = 1, linewidth = 2) +
  labs(x = "", y = "Study\nduration") +
  scale_x_continuous(transform = "log10",
                     limits = c(0.5,725),
                     breaks = c(1, 6, 12, 60, 
                                120, 240, 480), 
                     labels = c("1 month", "6 months", "1 year", "5 years", 
                                "10 years", "20 years", "40 years"),
                     expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), position = "right") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA),
        axis.text.x = element_text(size = 23, angle = 90,
                                   vjust = 0.5, hjust = 1),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_line(), 
        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
studyDurationPlot
ggsave(plot = studyDurationPlot, filename = "Figures/Duration.png",
       height = 3.5, width = 12)
```

```{r combine sites and duration}
#| fig-width: 12
#| fig-height: 6

combinedStudyScale <- siteQuantityPlot / studyDurationPlot
combinedStudyScale <- combinedStudyScale +
  theme(panel.background = element_rect(fill='transparent', color = NA),
        plot.background = element_rect(fill='transparent', color = NA))
combinedStudyScale
ggsave(plot = combinedStudyScale, filename = "Figures/StudyScale.png",
       height = 6, width = 12, bg='transparent')
```

## Covariates

#### Covariates table

#### Covariates figure

## Modelling

#### Modelling table

## Objectives

#### Objectives statistics

#### Objectives figure
