---
title: "Supporting information"
date: February 28 2025
abstract: "Supporting information for the article \"Twenty years of dynamic occupancy models: a review of applications and look to the future.\" This is divided into two sections: \"Review methods\", with additional details on how articles were processed, and \"Additional information\", with further figures and analyses which were not included in the published article. Session information from the R session used to run this document is provided at the end."
abstract-title: "Summary"
format:
  html:
    number-sections: true
    number-depth: 3
    theme: minty
embed-resources: true
---

```{r Packages_and_data}
#| include: false

# Packages
library(tidyverse)
library(readxl)

# Data from review spreadsheet pages
sheet_ArticleData <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                               sheet = "Article Data")
sheet_Objectives <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                              sheet = "Objectives") 
sheet_FocalTaxa <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                             sheet = "Focal Taxa") 
sheet_StudyArea <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                             sheet = "Study Area")
sheet_DataCollection <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                                  sheet = "Data Collection")
sheet_Covariates <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                              sheet = "Covariates")
sheet_Modelling <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                             sheet = "Modelling")
```

# Review methods

## Search queries

## Studies, datasets, and workflows

## Defining 'taxa'

## Covariates and parameters

# Additional information

## Authorship

### By country

### By objective

## Study locations and taxa

### By country

### Full taxa list

## Data collection

### Sampling method

### Site count by publication year

### Site count by study area scale

### Primary occasion count by secondary occasion count

### Primary occasion count by site count

### Dataset size (primary occasion count \* site count)

### Dataset size (primary occasion count \* secondary occasion count \* site count)

## Covariates

### Covariates table

### Covariate count by year

### Covariate count by site count

### Covariate count by primary occasion count

### Covariate count by study area scale

### Covariate count by dataset size (primary occasion count \* site count)

### Covariate count by objective

# Session information

```{r Session_Info}
#| echo: false
sessionInfo()
```